---
title: "Zarr-based spatial transcriptomics overview"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Zarr-based spatial transcriptomics overview}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

The Open Microscopy Environment "next-generation file format"
initiative [(OME-NGFF)](https://ngff.openmicroscopy.org/) has driven
substantial software development in the [scverse community](https://spatialdata.scverse.org/en/stable/).
This vignette describes R-based software being developed to convey
spatial omics solutions to investigators using NHGRI AnVIL.

# Package facts

We will use two packages being developed by a team assembled
for the 2024 hackathon in Basel, led by Helena Crowell.

```{r lkpacks,message=FALSE}
library(SpatialData)
library(SpatialData.plot)
library(SpatialData.data)
library(patchwork)
library(ggplot2)
library(sf)
utils::data(demo_spatialdata)
DT::datatable(demo_spatialdata)
```

# Example 1: MERFISH

Example datasets are retrieved from
cloud storage as needed and
cached locally.  For persistent
use, a `target` parameter can be used,
which will be populated with a Zarr store.

## Acquisition and persistence; folder layout

```{r getmer,message=FALSE}
mb = MouseBrainMERFISH()
mb
```

To continue working with the MERFISH archive,
set the `target` parameter in `MouseBrainMERFISH`
and then
use `readSpatialData`.


## Exhibiting the image

By default, the raw image is displayed at the coarsest
level of resolution.
```{r lkim}
plotSpatialData() + plotImage(mb)
```

## Cells and their coordinates

In this assay, cells are described by
a circle with given center and radius.
```{r lkloc}
shape(mb,2)@data |> sf::st_as_sf()
```

Coordinates for different components of the spatial data can require
transformation to be located or viewed together.

```{r lkcg}
plotCoordGraph(CTgraph(mb))
```

Here we see that there would need to be a transformation from
the rasterized image shown with plotImage, to bring it to the
global coordinates shared by all other modes.

## Visualizing the cells and annotation

There are two types of shape recorded in the data:
a partitioning of the plane into anatomical
layers, and (circular) segmentation of individual cells.
```{r dosh}
plotSpatialData() + plotShape(mb) + ggtitle("anatomy") | 
  plotSpatialData() + plotShape(mb) + 
     plotShape(mb,2, c="gray") + ggtitle("cells and anatomy")
```

## Transcripts and their locations

```{r lksh}
shape(mb,2)@data |> sf::st_as_sf()
```

## Quantifications of transcript abundance in cells

This experiment has measurements for 268 molecular
features.

```{r lkta2}
SpatialData::table(mb)
```

# Example 2: Breast cancer Xenium


## Acquisition

```{r lkbr1}
jbr1 = JanesickBreastXeniumRep1()
jbr1
```

## Cells and their coordinates

```{r lkbrc}
shape(jbr1)@data |> sf::st_as_sf()
```

The simple features representation lends itself
to rapid sketching of cell boundaries:

```{r lkxen}
cc = shape(jbr1)@data |> st_as_sf() |> tail(50000)
plotSpatialData() + geom_sf(data=cc)
```
